% !TEX root = report.tex
% !TEX program = pdflatex

During my internship, I developed a tool to solve counting constraints with arrays.
I also modified the model checker Sally to accept parametrized inputs, as well as
counting constraints. Sally can produce input for the counting constraints
solver and I was able to run bounded model checking and $k$-induction on a
simplied version of a parametric, fault-tolerant model of Byzantine General's
Problem.

Some work related to counting constraints had already been done. In \cite{AlbertiGP16, schweikardt}, there is already a way to solve
formulas with counting constraints. However, while they definitively explain that the problem
is decidable and why, they heavily rely on quantifier elimination and
thus may not be very practical nor easily integrated in modern SMT
solvers. I tried to compare the method of \cite{AlbertiGP16} and my solver,
unfortunately their implementation is not complete and most of our examples
cannot be checked by their tool.

The solver we developed can be improved. Divisibility constraints would be
useful to have a tool that supports all the Presburger arithmetic, even if we
did not find it useful for model checking related problems. Integration inside
an SMT solver could also provide a significant performance overhaul.

In the future, the decision procedure could be extended to support
interpolation and other advanced features of modern SMT solvers. Those features
are needed to implement modern and efficient model checking algorithms such as
IC3. The integration with existing solvers and the extensibility of the decision
procedure is the main contribution of my internship.

